---
title: "Predicting the Present - Unemployment"
author: "Gabrielle Martinez (Pace University)"
date: "6/25/2020"
output: html_document
---

```{r setup, include=FALSE}
library(scales)
library(broom)
library(modelr)
library(tidyverse)
library(here)
library(lubridate)

theme_set(theme_bw())
knitr::opts_chunk$set(echo = TRUE)
```

## Recreating the model from the paper using the source data

Initial unemployment stats from the Department of Labor: https://oui.doleta.gov/unemploy/claims.asp for the years 2004-2020

Load in data for cleaning and reformatting 

nsa = Not Seasonally Adjusted
sa = Seasonally Adjusted

We'll be starting off using Nonadjusted data (nsa)

```{r load_data}
unemployment <- read.csv("unemployment_initial.csv")

unemployment %>%
  select(Period, nsa, sa) %>%
  mutate(Period = as.Date(Period, format="%m/%d/%Y")) -> unemployment

unemployment %>%
  mutate(last_month_claims = lag(nsa, 1)) %>%
  filter(Period >= ' 2004-01-17' & Period <= '2011-07-02') %>%
  mutate(check = format(Period, "%Y-%m")) -> unemployment_2011

#load trends data
jobs <- read.csv("jobs_2011.csv")

welfare <- read.csv("welfare_2011.csv")

#reformat trends data
jobs %>%
  rename("jobs"="Geo..United.States") -> jobs

welfare %>%
  rename("welfare"="Geo..United.States") -> welfare


#seasonal adjustment of jobs trends data

ts(jobs$jobs, start = c(2004,1), frequency = 12) -> ts

stl(ts, s.window = "periodic") -> stl

stl[["time.series"]]-> ts

ts[,"trend"] -> ts

broom::tidy(ts) -> ts

ts %>%
  mutate(date = date_decimal(index))%>% 
  mutate(new_date = ceiling_date(date, "day")) %>% 
  mutate(period = format(new_date, "%Y-%m")) -> ts

inner_join(jobs, ts, by = c("Month" = "period")) %>% 
  select(Month, jobs, sa_jobs = value) -> jobs


#seasonal adjustment of welfare trends data

ts(welfare$welfare, start = c(2004,1), frequency = 12) -> ts

stl(ts, s.window = "periodic") -> stl

stl[["time.series"]]-> ts

ts[,"trend"] -> ts

broom::tidy(ts) -> ts

ts %>%
  mutate(date = date_decimal(index))%>% 
  mutate(new_date = ceiling_date(date, "day")) %>% 
  mutate(period = format(new_date, "%Y-%m")) -> ts

inner_join(welfare, ts, by = c("Month" = "period")) %>% 
  select(Month, welfare, sa_welfare = value) -> welfare


#join with trends data
left_join(unemployment_2011, jobs, by=c("check"="Month")) -> unemployment_2011
left_join(unemployment_2011, welfare, by=c("check"="Month")) -> unemployment_2011

unemployment_2011 %>%
  select(-check) -> unemployment_2011


```

Let's see what we're working with.

```{r}
ggplot(unemployment_2011) +
  geom_line(aes(y=log(sa), x=Period)) +
  ylab("log(claims)") +
  xlab("time")
```


### Remaking the in-sample models 

> "In this case, our baseline regression is a simple AR-1 model on the log of initial claims." [(Choi & Varian, 2011)](http://people.ischool.berkeley.edu/~hal/Papers/2011/ptp.pdf)

```{r model_nsa}
#unadjusted data
base_model <- lm(log(nsa) ~ log(last_month_claims), data = unemployment_2011)
tidy(base_model)
glance(base_model)

trend_model <- lm(log(nsa) ~ log(last_month_claims) + jobs + welfare, unemployment_2011)
tidy(trend_model)
glance(trend_model)
```

We get very different results from the paper when using the non-adjusted data to make our models. It maybe the case that we made a mistake when adjusting the trends data because the paper fails to describe how they adjusted the data beyond what function they used (`stl`). 

We know the authors used seasonally adjusted data and thus adjusted the trends data to match. Let's try that.

```{r seasonal_adjustment_model}

base_model_sa <- lm(log(sa) ~ log(last_month_claims), data = unemployment_2011)
tidy(base_model_sa)
glance(base_model_sa)

trend_model_sa <- lm(log(sa) ~ log(last_month_claims) + sa_jobs + sa_welfare, unemployment_2011)
tidy(trend_model_sa)
glance(trend_model_sa)

```
My R^2 are far smaller than the paper's but Basira had better luck than me in that department. While her coefficients much more closely align with the paper's, we both got intercepts that were much higher than the paper's. 

Let's see how these models look plotted

```{r}

unemployment_2011 %>% 
  add_predictions(base_model_sa, "base") %>%
  add_predictions(trend_model_sa, "trend") -> unemployment_2011_pred


ggplot(unemployment_2011_pred) +
  geom_line(aes(y=log(sa), x=Period)) +
  geom_line(aes(y=base,x= Period), color = "red", linetype = "dashed") +
  geom_line(aes(y=trend,x= Period), alpha = .5)
  

```



### Remaking Out-of-sample forecast model


```{r}


```







